### 2013-06-01 maize-ex.tex
WorkingWiki importation

\documentclass{article} 
\usepackage{fullpage} 
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\var}{\mathop{\mbox{Var}}}
\newcommand{\mutrate}{\lambda_{\mbox{mut}}}
\begin{document}

@@

\begin{table}[htb!!]
  \begin{center}
  \begin{tabular}{|crl|}
    \hline
    complete seed stock replacement prob & $p_e$     & 0.068 \\
    pollen migration rate               & $m_g$      & 0.0083 \\
    number of plants per field          & $N$        & $10^5$ \\
    number of ears used to replant      & $N_f$      & 561 \\
    mean ears per plant                 & $\mu_E$    & 3 \\
    partial stock replacement prob      & $p_m$      & 0.02 \\
    mean proportion stock replaced      & $r_m$      & 0.2 \\
    pollen migration distance           & $\sigma_p$ & 0 km \\
    seed replacement distance           & $\sigma_s$ & 50 km \\
    distance between demes              & $a$        & 15 km \\
    width of altitudinal cline          & $w$        & 62km \\
    deleterious selection coefficient   & $s_d$      & varies \\
    beneficial selection coefficient    & $s_b$      & varies \\
    slope of selection gradient         & $\alpha$   & $(s_d+s_b)/w$ \\
    variance in offspring number        & $\xi^2$    & varies \\
    maize population density            & $\rho$     & $5 \times 10^3$ \\
    \hline 
  \end{tabular}
\end{center}
  \caption{Parameter estimates used in calculations, and other notation.
  \label{tab:parameters}
  }
\end{table}

@@

@@

Since the probability of establishment at high altitude is approximately $2 s_b / \xi^2$, 
with $\xi^2$ the variance in offspring number,
the rate of appearance is just 
\begin{align*}
  \mutrate = 2 \rho A s_b / \xi^2 .
\end{align*}

@@

>>>>

\documentclass{article} 
\usepackage{fullpage} 
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\var}{\mathop{\mbox{Var}}}
\newcommand{\mutrate}{\lambda_{\mbox{mut}}}
\newcommand{\migrate}{\lambda_{\mbox{mig}}}
\begin{document}

@@

\begin{table}[htb!!]
  \begin{center}
  \begin{tabular}{|cll|}
    \hline
    complete seed stock replacement prob & $p_e$     & 0.068 \\
    pollen migration rate                & $m_g$      & 0.0083 \\
    number of plants per field           & $N$        & $10^5$ \\
    number of ears used to replant       & $N_f$      & 561 \\
    mean ears per plant                  & $\mu_E$    & 3 \\
    partial stock replacement prob       & $p_m$      & 0.02 \\
    mean proportion stock replaced       & $r_m$      & 0.2 \\
    pollen migration distance            & $\sigma_p$ & 0 km \\
    seed replacement distance            & $\sigma_s$ & 50 km \\
    distance between demes               & $a$        & 15 km \\
    width of altitudinal cline           & $w$        & 62km \\
    deleterious selection coefficient    & $s_d$      & varies \\
    beneficial selection coefficient     & $s_b$      & varies \\
    slope of selection gradient          & $\alpha$   & $(s_d+s_b)/w$ \\
    variance in offspring number         & $\xi^2$    & varies \\
    maize population density             & $\rho$     & $5 \times 10^3$ \\
    area of highland habitat             & $A$        & 500 km$^2$ \\
    mean dispersal distance              & $\sigma$   & 1.8 km \\
    \hline 
  \end{tabular}
\end{center}
  \caption{Parameter estimates used in calculations, and other notation.
  \label{tab:parameters}
  }
\end{table}

@@

\subsection{Dispersal distance}

The dispersal distance
-- the mean distance between parent and offspring --
is the average of the pollen and seed mean dispersal distances.
With the above assumptions, the pollen dispersal distance is zero,
and the seed dispersal distance is the chance of inter-village movement
multiplied by the mean distance moved.
This is
% pe <- .068; pm <- .02; rm <- .2; sigma.s <- 50
% (1/2) * (pe + (1-pe)*pm*rm) * sigma.s
\begin{align}
  \sigma = \frac{1}{2} (p_e + (1-p_e) p_m r_m ) \sigma_s = 1.7932 \text{km}
\end{align}
at the parameter values above.

@@

Since the probability of establishment at high altitude is approximately $2 s_b / \xi^2$, 
with $\xi^2$ the variance in offspring number,
the rate of appearance is just 
\begin{align*}
  \mutrate = 2 \rho A s_b \mu / \xi^2 .
\end{align*}
% rho <- 5000; sb <- 10^(-(1:3)); xisq <- 50; A <- 500
% sapply( c(1e-8,1e-5), function (mu) 2 * rho * A * sb *mu / xisq )
At the values above, with $.1 \le s_b \le .001$, the factor $2 \rho A s_b / \xi^2$ multiplying the mutation rate
varies between $10^2$ and $10^5$,
implying that a single-base mutation with $\mu=10^{-8}$ would have to wait between $10^4$ and $10^6$ generations to fix,
but a mutation with a larger target, say $\mu=10^{-5}$, would fix in tens to thousands of generations, depending on the selection coefficient.

@@

As we show in the theory paper, 
the rate of adaptation by diffusive migration is roughly
\begin{align*}
  \migrate = \rho \frac{s_b \sqrt{2 s_m}}{2\xi^2} \exp\left(- \frac{\sqrt{2 s_m}R}{\sigma} \right) .
\end{align*}
We can talk about this more and give some simulations.
But for now, let's interpret.

First note that for $10^{-1} \le s_m \le 10^{-4}$, the value $1/\sqrt{2s_m}$ is between 2 and 70 --
so the exponential decay of the chance of migration fallws off on a scale of between 2 and 70 times the dispersal distance.
Above we have estimated the dispersal distance to be $\sigma \approx 2$ km,
and far below the mean distance $\sigma_s$ to the field that a farmer replants seed from, when this happens,
which we have as $\sigma_s = 50$ km.
Taking $\sigma=2$ km, we have that $4 \le \sigma/\sqrt{2s_m} \le 150$ km.
A very conservative upper bound might be $\sigma \le \sigma_s/20$ (if farmers replaced 10\% of their seed with long-distance seed every year).
At this upper bound, we would have $5 \le \sigma/\sqrt{2s_m} \le 175$ km,
which is not very different.
This makes the exponential term very small since $R$ is on the order of 1,000 km.

Taking $\sigma=2$ km, we then compute that 
% rho <- 5000; sb <- .01; sm <- 10^(-(1:4)); xisq <- 50; sigma <- 2
% sapply( 1000*(1:4), function (R) rho * sb * sqrt(2*sm) / (2 *  xisq) * exp(- sqrt(2*sm)*R/sigma ) )
if $s_m = 10^{-4}$ (very weak selection in the lowlands), then for $R=1,000$ km, the migration rate is $\migrate \le 10^{-5}$,
i.e.\ it would take on the order of 100,000 generations (years) to get a successful migrant only 1,000 km away,
under this model of undirected, diffusive dispersal.
For larger $s_m$, the migration rate is much smaller.

%%%%%%%%%%%%
\section{Conclusion}

It seems unlikely that any alleles that are adaptive in the highlands and deleterious at all in the lowlands
would have transited central America by undirected (diffusive) sharing of seed.
The conclusions could change if we drastically underestimate the rate of very long distance sharing of seed,
e.g.\ if sharing across hundreds of kilometers was common at some point.

Both calculations are very pessimistic about the chance of shared single-base changes through either migration or independent mutation.
However, independent mutations could be expected in kilobase-size targets,
suggesting there might be signal for genes that share adaptive changes.
